<<<<<<< HEAD
---
title: 'Jonathan, Environmental Data Science @UIUC. Lab 9 - Watershed analysis'
date: "3/25/2024"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Overview

In this lab, you will practice watershed analysis in R using a DEM file downloaded by yourself.

As a preparation, please go through the exercise under Canvas module > Week 11 > Watershed-analysis-tutorial-updated.html/rmd, and understand the purpose and meaning of each script line. 

Then, replicate the entire watershed analysis process as described in Watershed-analysis-tutorial-updated.html using a DEM tile covering your own region of interest (e.g., part of the Rocky Mountain, Flint Hills, or Central Illinois) in the US. In other words, you will need to download your own DEM, instead of using the DEM available in the spDataLarge package.

One platform to download high-resolution national DEM is The National Map (TNM) Download Apps: https://apps.nationalmap.gov/downloader. 

Once you are on the TNM webpage, zoom to your area of interest using the interactive map, and then further refine your region using the "Extent", "Polygon", or "Point" option after "Area of Interest". 

Then, check "Elevation Products (3DEP)" and select the DEM with your preferred spatial resolution (e.g., 1/3 arc-second, which is about ~10 m). Think about what resolution will serve your purpose (note: very high-resolution DEM is not always necessary). 

Under "File Formats", check the GeoTIFF format (.tif).

After the region is defined and the DEM and the file format are specified, click on "Search Products". The DEM tiles satisfying your search criteria will be returned. If multiple files are returned, you will need to download only ONE of them for the purpose of this lab. You can click on "Footprint" to see where the DEM tile covers. Once you have decided which tile to download, click on "Download Link (TIF)", and the DEM tile will be downloaded to your computer. 

Now you have your own DEM! Congrats! And you are ready to continue following the procedure described in Watershed-analysis-tutorial-updated.html. 

Submission: A fully commented and formatted R Markdown file that replicates each of the steps in the entire watershed analysis as described in Watershed-analysis-tutorial-updated.rmd but based on your downloaded DEM.

To start: load the packages and define the variable environments as below...

```{r setup-libraries}
#install.packages("rgl")
library(tidyverse)
library(raster)
library(sf)
library(tmap)
library(stars)
library(rgl)
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)

#install.packages("rayshader")
library(rayshader) 

#install.packages("whitebox", repos = "http://R-Forge.R-project.org")
#install.packages("whitebox")
whitebox::install_whitebox()
library(whitebox)
```

```{r environment-setup, warning=FALSE, message=FALSE}

# Area of interest -> missouri river area
# Set working directory to location of DEM.tif file
my_directory <- "/Users/jonathanbernard/Desktop/UIUC/Spring 2024/GGIS 224/GGIS224/Labs/Lab9"
setwd(my_directory)

# Read the dem
dem = rast("dem.tif")

crs(dem)

tmap_mode("view")


if (ncol(dem)<1000 & nrow(dem)<1000)
{
  tm_shape(dem)+
    tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE)+
    tm_scale_bar()
} else {
  tm_shape(aggregate(dem, 10), raster.downsample = TRUE)+
    tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE)+
    tm_scale_bar()
}

## map hillshade
wbt_hillshade(dem = paste0(my_directory, "/dem.tif"),  output = paste0(my_directory, "/hillshade.tif"), azimuth = 300) 
#Read the produced hillshade 
hillshade = raster("hillshade.tif") 

# Map the hillshade
if (ncol(hillshade)<1000 & nrow(hillshade)<1000)
{
  tm_shape(hillshade)+
    tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
    tm_scale_bar()
} else {
  tm_shape(aggregate(hillshade, 10), raster.downsample = TRUE)+
    tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
    tm_scale_bar()
}

## Prepare DEM for Hydrology Analyses
#fill the DEM
wbt_fill_single_cell_pits(dem = paste0(my_directory, "/dem.tif"),  output = paste0(my_directory, "/dem_filled.tif"))
wbt_breach_depressions_least_cost(
  dem = paste0(my_directory, "/dem_filled.tif"),
  output = paste0(my_directory, "/dem_breached.tif"),
  dist = 5,
  fill = TRUE) 
wbt_fill_depressions_wang_and_liu(
  dem = paste0(my_directory, "/dem_breached.tif"),
  output = paste0(my_directory, "/dem_filled_breached.tif"),
)

#Read the smoothed dem
dem = rast("dem_filled_breached.tif") 

#Map the DEM
if (ncol(dem)<1000 & nrow(dem)<1000)
{
  tm_shape(dem)+
    tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE)+
    tm_scale_bar()
} else {
  tm_shape(aggregate(dem, 10), raster.downsample = TRUE)+
    tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE)+
    tm_scale_bar()
}

## Create flow accumulation and pointer grids (flow direction)
#Generate the flow accumulation map. Make sure to use the filled DEM "dem_filled_breached.tif"!!
wbt_d8_flow_accumulation(input = paste0(my_directory, "/dem_filled_breached.tif"), output = paste0(my_directory, "/D8FA.tif"))
#Read the produced flow accumulation file
wbt_d8_flow_accumulation_lyr = rast("D8FA.tif") 

#Visualize the flow accumulation map
if (ncol(wbt_d8_flow_accumulation_lyr)<1000 & nrow(wbt_d8_flow_accumulation_lyr)<1000)
{
  tm_shape(wbt_d8_flow_accumulation_lyr)+
    tm_raster(style = "quantile", legend.show = TRUE)+ #Fisher or quantile classification method is used to enhance visualization for flow accumulation (distribution dominated by small values)
    tm_scale_bar()
} else {
  tm_shape(aggregate(wbt_d8_flow_accumulation_lyr, 10), raster.downsample = TRUE)+
    tm_raster(style = "quantile", legend.show = TRUE)+ #Fisher or quantile classification method is used to enhance visualization for flow accumulation (distribution dominated by small values)
    tm_scale_bar()
}

#Generate the flow direction (pointer) map
wbt_d8_pointer(dem = paste0(my_directory, "/dem_filled_breached.tif"), output = paste0(my_directory, "/D8pointer.tif"))

#Read the produced flow direction file
wbt_d8_pointer_lyr = rast("D8pointer.tif")

#Visualize the flow direction map
if (ncol(wbt_d8_pointer_lyr)<1000 & nrow(wbt_d8_pointer_lyr)<1000)
{
  tm_shape(wbt_d8_pointer_lyr)+
    tm_raster(style = "cont", legend.show = TRUE)+
    tm_scale_bar()
} else {
  tm_shape(aggregate(wbt_d8_pointer_lyr, 10, fun = modal), raster.downsample = TRUE)+ #here modal function is used for aggregating the discrete D8 flow directions.
    tm_raster(style = "cont", legend.show = TRUE)+
    tm_scale_bar()
}

## Set pour points
ppoints <- tribble(
  ~Lon, ~Lat,
  -92.5495, 38.2211,
  -92.1952, 38.4127
)

#Create dataframe with pour points (make sure to set up the correct spatial coordinate system)
ppointsSP <- SpatialPoints(ppoints, proj4string = CRS("+proj=longlat +datum=WGS84")) #

#Convert data frame to shapefile and write the shapefile to our data directory
shapefile(ppointsSP, filename = "pourpoints.shp", overwrite = TRUE)

#Read the shapefile
ppointsSP_gcs = st_read("pourpoints.shp")

st_crs(dem) #4326

ppointsSP_prj = st_transform(ppointsSP_gcs, crs="EPSG:4326")

st_write(ppointsSP_prj, "ppointsSP_prj.shp", layer_options = 'OVERWRITE=YES', update = TRUE)

wbt_extract_streams(flow_accum = paste0(my_directory, "/D8FA.tif"),
                    output = paste0(my_directory, "/raster_streams.tif"),
                    threshold = 50)

wbt_jenson_snap_pour_points(
  pour_pts = paste0(my_directory, "/ppointsSP_prj.shp"),
  streams = paste0(my_directory, "/raster_streams.tif"),
  output = paste0(my_directory, "/snappedpp.shp"),
  snap_dist = 10
)

# Read the produced files
streams <- raster("raster_streams.tif") # Read the generated stream network
pp <- shapefile("snappedpp.shp") # Read the snapped pour points

#Visualize the pour points BEFORE being snapped to the network. 
if (ncol(streams)<1000 & nrow(streams)<1000)
{
  tm_shape(streams)+
    tm_raster(legend.show = TRUE, palette = "Blues")+
    tm_shape(ppointsSP_prj)+
    tm_dots(col = "red")
} else {
  tm_shape(aggregate(streams, 10, fun = max), raster.downsample = TRUE)+ #downsample function set to be MAX to improve continuity of the stream network. 
    tm_raster(legend.show = TRUE, palette = "Blues")+
    tm_shape(ppointsSP_prj)+
    tm_dots(col = "red")
}

#Visualize the pour points AFTER being snapped to the network. 
if (ncol(streams)<1000 & nrow(streams)<1000)
{
  tm_shape(streams)+
    tm_raster(legend.show = TRUE, palette = "Blues")+
    tm_shape(pp)+
    tm_dots(col = "red")
} else {
  tm_shape(aggregate(streams, 10, fun = max), raster.downsample = TRUE)+ #downsample function set to be MAX to improve continuity of the stream network. 
    tm_raster(legend.show = TRUE, palette = "Blues")+
    tm_shape(pp)+
    tm_dots(col = "red")
}

## Delineate watersheds
wbt_watershed(d8_pntr = paste0(my_directory,"/D8pointer.tif"),
              pour_pts = paste0(my_directory,"/snappedpp.shp"),
              output = paste0(my_directory,"/watersheds.tif"))
#Read the produced watershed file
ws = raster("watersheds.tif") #Read the watershed raster file

#Visualize the watersheds
if (ncol(hillshade)<1000 & nrow(hillshade)<1000)
{
  tm_shape(hillshade)+
    tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
    tm_shape(ws)+
    tm_raster(legend.show = TRUE, alpha = 0.5, style = "cat")+
    tm_shape(pp)+
    tm_dots(col = "red")
} else {
  tm_shape(aggregate(hillshade, 10), raster.downsample = TRUE)+
    tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
    tm_shape(aggregate(ws, 10, fun = max), raster.downsample = TRUE)+
    tm_raster(legend.show = TRUE, alpha = 0.5, style = "cat")+
    tm_shape(pp)+
    tm_dots(col = "red")
}

## Convert watersheds to shapefiles
#Convert watershed to shapefile
wsshape = st_as_stars(ws) %>% st_as_sf(merge = T)
wsshape

wsshape_dissolved <- wsshape %>% 
  group_by(watersheds) %>% 
  summarise()

st_write(wsshape_dissolved, 'watersheds_shape.shp', layer_options = 'OVERWRITE=YES', update = TRUE)

#Visualize the vectorized watersheds
if (ncol(hillshade)<1000 & nrow(hillshade)<1000)
{
  tm_shape(hillshade)+
    tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
    tm_shape(wsshape_dissolved)+
    tm_borders(col = "red")
} else {
  tm_shape(aggregate(hillshade, 10), raster.downsample = TRUE)+
    tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
    tm_shape(wsshape_dissolved)+
    tm_borders(col = "red")
}

#Extract DEM pixels within each watershed
wsElevs <- extract(dem, wsshape_dissolved)

WS1 = filter(wsElevs, ID == 1) #Pixels in watershed 1
WS2 = filter(wsElevs, ID == 2) #Pixels in watershed 2

ggplot() +
  geom_density(as_tibble(WS1[,2]), mapping = aes(value, color = "WS1"))+
  geom_density(as_tibble(WS2[,2]), mapping = aes(value, color = "WS2"))+
  xlab("Elevation (meters above sea level)")
```
=======
---
title: 'xxx, Environmental Data Science @UIUC. Lab 9 - Watershed analysis'
date: "3/25/2024"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Overview

In this lab, you will practice watershed analysis in R using a DEM file downloaded by yourself.

As a preparation, please go through the exercise under Canvas module > Week 11 > Watershed-analysis-tutorial-updated.html/rmd, and understand the purpose and meaning of each script line. 

Then, replicate the entire watershed analysis process as described in Watershed-analysis-tutorial-updated.html using a DEM tile covering your own region of interest (e.g., part of the Rocky Mountain, Flint Hills, or Central Illinois) in the US. In other words, you will need to download your own DEM, instead of using the DEM available in the spDataLarge package.

One platform to download high-resolution national DEM is The National Map (TNM) Download Apps: https://apps.nationalmap.gov/downloader. 

Once you are on the TNM webpage, zoom to your area of interest using the interactive map, and then further refine your region using the "Extent", "Polygon", or "Point" option after "Area of Interest". 

Then, check "Elevation Products (3DEP)" and select the DEM with your preferred spatial resolution (e.g., 1/3 arc-second, which is about ~10 m). Think about what resolution will serve your purpose (note: very high-resolution DEM is not always necessary). 

Under "File Formats", check the GeoTIFF format (.tif).

After the region is defined and the DEM and the file format are specified, click on "Search Products". The DEM tiles satisfying your search criteria will be returned. If multiple files are returned, you will need to download only ONE of them for the purpose of this lab. You can click on "Footprint" to see where the DEM tile covers. Once you have decided which tile to download, click on "Download Link (TIF)", and the DEM tile will be downloaded to your computer. 

Now you have your own DEM! Congrats! And you are ready to continue following the procedure described in Watershed-analysis-tutorial-updated.html. 

Submission: A fully commented and formatted R Markdown file that replicates each of the steps in the entire watershed analysis as described in Watershed-analysis-tutorial-updated.rmd but based on your downloaded DEM.

To start: load the packages and define the variable environments as below...

```{r environment-setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(raster)
library(sf)
library(tmap)
library(stars)
library(rgl)
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)

#install.packages("rayshader")
library(rayshader) 

#install.packages("whitebox", repos="http://R-Forge.R-project.org")
library(whitebox)
whitebox::wbt_init()
```
>>>>>>> 50e4cc1cdd4edf72ae235014b61b99d93e7dfcfa
