---
title: 'GGIS 224 - Lab 7'
author: 'Name, Environmental Data Science @UIUC '
date: "2/xx/2024"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This lab is dedicated to raster analysis and raster-vector interactions. 

In our previous lectures, We emphasized attribute operations, spatial operations, and geometric operations primarily using vector data. The same concepts and operations also apply to raster data. Given the expectation that you have mastered the materials for vector data, you should be able to self-study [Section 4.3. Spatial operations on raster data](https://r.geocompx.org/spatial-operations), [Section 5.3. Geometric operations on raster data](https://r.geocompx.org/geometry-operations), and [Chapter 6. Raster-vector interactions](https://r.geocompx.org/raster-vector) and learn how the concepts and operations can be applied to raster data and how both data types can be used interactively. 

After you have gone through Section 4.3, Section 5.3, and Chapter 6, explore the following interesting questions. 

Don't forget to comment your code while answering the questions below.    

Render your completed RmD document by clicking on the "Knit" option in RStudio. Upload the .html file to your lab submission.
Your uploaded .html file should follow this naming convention: Lab7_FirstNameInitial_LastName.html (for example, Lab7_E_Hall.html).


---

## Environment Setup

Load up all libraries needed for this lab. 

```{r environment-setup, warning=FALSE, message=FALSE}
library(sf)
library(terra)
library(dplyr)
library(spData)
```

---

Some of the following exercises use a vector (zion_points) and raster dataset (srtm) from the spDataLarge package. They also use a polygonal ‘convex hull’ derived from the vector dataset (ch) to represent the area of interest.

---

```{r code chunk for loading dataset}
zion_points_path = system.file("vector/zion_points.gpkg", package = "spDataLarge")
zion_points = read_sf(zion_points_path)
srtm = rast(system.file("raster/srtm.tif", package = "spDataLarge"))
ch = st_combine(zion_points) |>
  st_convex_hull() |> 
  st_as_sf()

# Plot the data to understand their geography
plot(srtm)
plot(zion_points, add =TRUE)
plot(ch, add = TRUE)
```


---


## Explore and answer the following questions. 

Q1.1. Crop the srtm raster using (1) the zion_points dataset and (2) the ch dataset. 
Are there any differences in the output maps? 


```{r code chunk for Q1.1}
#Enter your code chunk that explores this question. Please leave comments as in detail as possible to show your logical development. 
```

Are there any differences in the cropped maps? Why? 

*ADD YOUR ANSWER HERE: *


Q1.2. Instead of cropping, mask srtm with zion_points and ch. Are there any differences in the output maps? If so, why?


```{r code chunk for Q1.2}
#Enter your code chunk that explores this question. Please leave comments as in detail as possible to show your logical development. 
```

Are there any differences in the masked maps? Why?

*ADD YOUR ANSWER HERE: *


---


Q2. Extract elevation values:
(1) Extract values from the srtm raster with zion_points. 
(2) Next, extract average (mean) values of srtm using a 500-meter buffer around each point (in zion_points).
(3) Compare these two sets of values in the same layer (dataframe). 


```{r code chunk for Q2}
#Enter your code chunk that explores this question. Please leave comments as in detail as possible to show your logical development. 
```

When would extracting values by buffers be more suitable than by points alone?

*ADD YOUR ANSWER HERE: Open question. Feel free to discuss anything based on your perspectives (as long as the answer is reasonable). *


---


Q3. Rasterize elevation points:
(1) Subset points higher than 3100 meters in New Zealand (the nz_height object).
(2) Create a template raster with a resolution of 3 km for the extent of the new point dataset.
(3) Using these two new objects, rasterize the points to display the number of points>3100m (count) in each grid cell. Plot the result.
(4) Now rasterize the poings to display the maximum elevation point in each 3km grid cell. Plot the result. 


```{r code chunk for Q3}
#Enter your code chunk that explores this question. Please leave comments as in detail as possible to show your logical development. 
```


---


Q4. Raster Local operations (4.3.3) and Zonal operations (4.3.5):
(1) Using local operations, use *dem = rast(system.file("raster/dem.tif", package = "spDataLarge"))* and reclassify the elevation in three classes: low (<300), medium (300-500), and high (>500).
(2) Then with zonal operations, read the NDVI raster (ndvi = rast(system.file("raster/ndvi.tif", package = "spDataLarge"))) and compute the mean NDVI *and* the mean elevation for each altitudinal class.


```{r code chunk for Q4}
#Enter your code chunk that explores this question. Please leave comments as in detail as possible to show your logical development. 
```

---


Q5. Use the Landsat image provided by the spDataLarge package: *rast(system.file("raster/landsat.tif", package = "spDataLarge"))* to calculate the Normalized Difference Water Index (NDWI; (green - nir)/(green + nir)) of the image. NDWI is similar to NDVI (see Section 4.3.3) except that NDVI is to enhance the appearance of vegetation from a multispectral image whereas NDWI is to enhance the appearance of water.  

Hint: For NDVI and NDWI calculation, please follow the procedure described in Section 4.3.3 of Chapter 4. 
As described in Chapter 4, the image (which is a Landsat level-2 product) has four spectral bands from the Landsat 8 satellite — blue (band 1 in the image), green (band 2), red (band 3), and near-infrared (NIR, band 4). The Landsat level-2 products are stored as integers to save disk space, and thus we need to convert them to floating-point numbers before doing any calculations. For that purpose, we need to apply a scaling factor (0.0000275) and add an offset (-0.2) to the original values

Bonus: 1 point. Calculate a correlation between NDVI and NDWI for this area. You can use the [layerCor()](https://rspatial.github.io/terra/reference/layerCor.html) function, and refer to Section 3.3.1 in Chapter 3 for stacking multilayered rasters.


```{r code chunk for Q5}
#Enter your code chunk that explores this question. Please leave comments as in detail as possible to show your logical development. 
```

---

Render your document by clicking on the "Knit" option in RStudio. Upload the .html file to your assignment submission!

.